---
title: "Traits and trees"
author: "Rutger Vos (@rvosa)"
date: "1-11-2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analyzing trees and traits

In this example we are going to explore whether domesticated ungulates are closer to each 
other in trait values than they are in evolutionary relatedness. Only some ungulates have 
ever been domesticated, and there are indications that the succesful ones have been 
domesticated multiple times (cows, pigs), whereas others have been failures. Apparently 
there are combinations of trait values among ungulates that determine whether domestication 
can ever succeed, and only some of them have the right combination (Jared Diamond's Anna 
Karenina argument). Thus, average pairwise distance among domesticated ungulates clustered 
by Gower distance should be shorter than everage pairwise evolutionary distance.

To assess this, we are going to build on the [occurrences/trees](occurrences-trees.Rmd) 
tutorial. First, we need a tree for the Ungulates.

```{r make_phylo}
library(taxize)
library(DBI)
library(ape)
source('../R/expand_taxon.R')

# 'ungulates' is a common name, not an accepted higher taxon.
# So what accepted taxa fall under this name?
ungulates <- unique(comm2sci("Ungulates",db="itis")$Ungulates)

# locations of data files
db_file <- '../data/sql/tgd.db'
msw3_file <- '../data/taxa/msw3-all.csv'
supertree_file <- '../data/phylogeny/Bininda-emonds_2007_mammals.tsv'

# we need the tree_id of the msw3_file
db <- dbConnect(RSQLite::SQLite(), db_file)
query <- 'select tree_id from trees where tree_name="%s"'
tree_id <- dbGetQuery(db, sprintf(query, msw3_file))$tree_id

# iterate over the top level 'ungulates' and expand them to species
names <- list()
for ( name in ungulates ) {
	
	# the actual call to the file we sourced above
	expanded <- expand.taxon(
		db_file     = db_file,
		taxon_name  = toupper(name),
		db          = db,
		tree_id     = tree_id,
		taxon_level = 'SPECIES'
	)
	
	# post process to get a flat list
	nnames <- length(expanded)
	expanded_names <- vector(mode="list", length=nnames)
	for ( i in 1:nnames ) {
		expanded_names[[i]] <- expanded[[i]]$taxon_name
	}
	names <- c(names,expanded_names)
}

# now switch to the mammal supertree and extract its subtree
source('../R/make_phylo.R')
tree_id <- dbGetQuery(db, sprintf(query, supertree_file))$tree_id
tree <- make.phylo(db_file, tree_id, names)
plot(tree, show.tip.label=F)
axisPhylo(side = 1)
```
Now that we have a tree, we can look for character state values for the tips in 
in the tree. As follows:
```{r make_traitset}
source('../R/make_traitset.R')

# we use the traits from PanTHERIA
df <- make.traitset(db_file, tree$tip.label, 'PanTHERIA_1-0_WR05_Aug2008')
```

Now let's do an analysis:

```{r gowdis}
library(FD)

# in this case all values are numeric, but
# come out of the database as character, 
# so we coerce them here
for (i in 1:50) { 
	df[,i] <- as.numeric(df[,i])
}

# here we compute the Gower distance from FD,
# then compute a dendrogram using neighbor-joining
dist <- gowdis(df)
plot(njs(dist), type = "unrooted")
```
